The Constant-Q Transform Spectral Envelope Coefficients: A Timbre Feature Designed for Music

### 2021-07-14 IEEE SP Magazine/07-2021 - Draft/article.tex
Finished article!

\IEEEPARstart{T}{imbre} is the attribute of sound which makes, for example, two musical instruments playing the same note sound different. It is typically associated with the spectral (but also the temporal) envelope and assumed to be independent from the pitch (but also the loudness) of the sound \cite{moore2004}. In this article, we will show how to design a simple but functional pitch-independent timbre feature which is well adapted to musical data, by deriving it from the constant-Q transform (CQT), a log-frequency transform which matches the equal tempered musical scale \cite{brown1991, brown1992}. We will show how to decompose the CQT spectrum into an energy-normalized pitch component and a pitch-independent spectral envelope, the latter from which we will extract a number of timbral coefficients. We will then evaluate the discriminative power of these \textit{CQT spectral envelope coefficients (CQT-SEC)} on the NSynth dataset \cite{engel2017}, a large-scale dataset of musical notes which is publicly available, comparing them with the mel-frequency cepstral coefficients (MFCCs) \cite{mermelstein1976}, features originally designed for speech recognition but commonly used to characterize timbre in music.

@@

A timbre feature which is well adapted to musical data, pitch-independent, and with high discriminative power can find uses in a number of applications, such as similarity detection, sound recognition, and audio classification, in particular, of musical instruments. Additionally, the ability to decompose the spectrum of a sound (here, the CQT spectrum) into a pitch-independent spectral envelope and an energy-normalized pitch component can be useful for analysis, transformation, and resynthesis of music signals. The energy-normalized pitch component can also potentially be used for tasks such as pitch identification, melody extraction, and chord recognition.

@@

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one simple characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound, which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when they need one simple timbre feature \cite{mueller2007}. While the MFCCs were shown to be practical in some MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily well adapted to musical data. In particular, they are derived through an old process which makes use of the mel scale, a perceptual scale experimentally designed 85 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but typically in terms of implicit embeddings which are tied to a specific trained model \cite{engel2017}, \cite{pons2017}, \cite{hung2019}, \cite{luo2019}, \cite{lee2020} and not necessarily as explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to feature to characterize timbre by MIR practitioners.

@@

We present here the CQT-SECs, a novel timbre feature which is well adapted to musical data, pitch-independent, simple to compute, interpretable, and functional. We will show how to derive it from the CQT, a frequency transform with a logarithmic resolution which matches the notes of the equal temperament, a musical scale typically used in Western music \cite{brown1991, brown1992}. We will first show how to decompose the CQT spectrum into a pitch-independent spectral envelope and an energy-normalized pitch component, and then extract a number of timbral coefficients from the spectral envelope.

@@

We start with the assumption that a log spectrum $S$, in particular, the CQT spectrum, can be represented as the convolution between a pitch-independent spectral envelope $E$ (which mostly contains the timbre information) and an energy-normalized pitch component $P$ (which mostly contains the pitch information), as shown in Equation \ref{eq:assumption}, where $*$ represents the convolution operation. 
\begin{equation}
\label{eq:assumption}
S = E * P
\end{equation}

This convolution process can be thought of as a source-filter model \cite{fant1970} which is here not applied in the time domain but in the frequency domain, with the source and the filter being the pitch component and the envelope, respectively.

@@

Figure \ref{fig:deconvolution} shows an example of deconvolution of a CQT spectrogram into its envelope and pitch component. The CQT spectogram was computed from an audio signal created by concatenating 12 4-second notes of an acoustic bass playing from C1 (32.70 Hz) to B1 (61.74 Hz) in ascending order. The notes come from the NSynth dataset \cite{engel2017} and correspond to instrument id \texttt{bass\_acoustic\_000}, MIDI numbers \texttt{024} to \texttt{035}, and velocity number \texttt{075}. The CQT spectrogram was computed using librosa \cite{mcfee2015,schoerkhuber2010}, with a sampling rate of 16 kHz, a hop length of 512 samples, a minimum frequency of 32.70 Hz (corresponding to C1), 95 frequency bins, and 12 bins per octave. As we can see, the envelope looks as if the CQT spectrogram has been normalized in pitch, with all the notes being brought down to the lowest frequency (corresponding to C1); while the pitch component looks as if the CQT spectrogram has been stripped down from all its energy, leaving mostly the fundamental frequencies of the notes. Note that in practice, we use a power CQT spectrogram (i.e., magnitude to the power of 2) and take the real part of the envelope and pitch component to ensure real values. We can also potentially post-process this deconvolution, for example, by zeroing the few negative values in the pitch component and using it to derive a refined  envelope from the log spectrum.

@@

Figure \ref{fig:extraction} shows an example of CQT-SECs, on the left plot. 20 coefficients were extracted from the envelope resulting from the deconvolution of the CQT spectrogram of the musical signal shown in Figure \ref{fig:deconvolution}. These coefficients essentially correspond to the harmonics of the musical instrument which contain most of its spectral energy and can therefore be a reasonable representation of its timbre. For comparison, we also show the MFCCs computed from the same musical signal, on the right plot. 20 coefficients were computed using librosa \cite{mcfee2015}, with a sampling rate of 16 kHz, a window length of 1024 samples, and hop length of 512 samples, matching the time resolution of the CQT-SECs so that both features have the same size in time and frequency.

\begin{figure*}[htp]
    \centering
    \includegraphics[width=\textwidth]{extraction.png}
    \caption{CQT-SECs extracted from the envelope obtained following the deconvolution shown in Figure \ref{fig:deconvolution} (left plot, shown in dB) and MFCCs computed from the same musical signal (right plot).}
    \label{fig:extraction}
\end{figure*}

@@

We derived the CQT-SECs for all the notes in this subset, by first computing the CQT spectrogram using librosa, with a sampling rate of 16 kHz, a hop length of 512 samples, a minimum frequency of 32.70 Hz (corresponding to C1), 95 frequency bins, and 12 bins per octave, and then extracting 20 coefficients from the envelope resulting from the deconvolution of the CQT spectrogram, leading to CQT-SECs of size 20 by 126 (time frames). We used a power CQT spectrogram (i.e., magnitude to the power of 2) and took the real part of the envelope to ensure real values. For comparison, we also computed the MFCCs, using librosa, with a sampling rate of 16 kHz, a window length of 1024 samples, a hop length of 512 samples, and 20 coefficients, leading to MFCCs of same size as the CQT-SECs, i.e., 20 by 126.

@@

We also averaged the note similarities over every one of the 11 instrument families, leading to similarity matrices of size 11 by 11, for both the CQT-SECs and the MFCCs. Figure \ref{fig:family_similarities} shows these instrument family similarity matrices for the CQT-SECs and the MFCCs. As we can see here as well, the similarities for the CQT-SECs have more variance, showing mostly clearer high self-similarities and low cross-similarities, while the similarities for the MFCCs are mostly very high. Figure \ref{fig:family_similarities2} additionally shows the self-similarities (in green) and the error bars for the cross-similarities (means and standard deviations in red) for every instrument family, for the CQT-SECs and the MFCCs. As we can again see, the self-similarities for the CQT-SECs are noticeably and mostly higher than the means plus standard deviations (except for the string and synth\_lead families), while the self-similarities and cross-similarities for the MFCCs are all very high and harder to tell apart.

@@

We have shown that we can derive a simple but function timbre feature which is more adapted to musical data ...

%Limitations:
%- just like MFCCs, not very robust to noise, to mixture of musical instruments playing at different pitches

>>>>

\IEEEPARstart{T}{imbre} is the attribute of sound which makes, for example, two musical instruments playing the same note sound different. It is typically associated with the spectral (but also the temporal) envelope and assumed to be independent from the pitch (but also the loudness) of the sound \cite{moore2004}. In this article, we will show how to design a simple but functional pitch-independent timbre feature which is well adapted to musical data, by deriving it from the constant-Q transform (CQT), a log-frequency transform which matches the equal-tempered musical scale \cite{brown1991, brown1992}. We will show how to decompose the CQT spectrum into an energy-normalized pitch component and a pitch-independent spectral envelope, the latter from which we will extract a number of timbral coefficients. We will then evaluate the discriminative power of these \textit{CQT spectral envelope coefficients (CQT-SEC)} on the NSynth dataset \cite{engel2017}, a large-scale dataset of musical notes which is publicly available, comparing them with the mel-frequency cepstral coefficients (MFCCs) \cite{mermelstein1976}, features originally designed for speech recognition but commonly used to characterize timbre in music.

@@

A timbre feature which is well adapted to musical data, pitch-independent, and with high discriminative power can find uses in a number of applications, such as similarity detection, sound recognition, and audio classification, in particular, of musical instruments. Additionally, the ability to decompose the spectrum of a sound (here, the CQT spectrum) into a pitch-independent spectral envelope and an energy-normalized pitch component can be useful for analysis, transformation, and resynthesis of music signals. The energy-normalized pitch component can also potentially be used for tasks such as pitch identification and melody extraction.

@@

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one simple characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound, which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when they need one simple timbre feature \cite{mueller2007}. While the MFCCs were shown to be practical in some MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily well adapted to musical data. In particular, they are derived through an old process which makes use of the mel scale, a perceptual scale experimentally designed 85 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but typically in terms of implicit embeddings which are tied to specific trained models \cite{engel2017, pons2017, hung2019, luo2019, lee2020} and not necessarily as explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to feature to characterize timbre by MIR practitioners.

@@

We hereby present the CQT-SECs, a novel timbre feature which is well adapted to musical data, pitch-independent, simple to compute, interpretable, and functional. We will show how to derive it from the CQT, a frequency transform with a logarithmic resolution which matches the notes of the equal temperament, a tuning system typically used in Western music \cite{brown1991, brown1992}. We will first show how to decompose the CQT spectrum into a pitch-independent spectral envelope and an energy-normalized pitch component, and then extract a number of timbral coefficients from the spectral envelope.

@@

We start with the assumption that a log spectrum $S$, in particular, the CQT spectrum, can be represented as the convolution between a pitch-independent spectral envelope $E$ (which mostly contains the timbre information) and an energy-normalized pitch component $P$ (which mostly contains the pitch information), as shown in Equation \ref{eq:assumption}, where $*$ represents the convolution operation. 
\begin{equation}
\label{eq:assumption}
S = E * P
\end{equation}
This convolution process can be thought of as a source-filter model \cite{fant1970} which is here not applied in the time domain but in the frequency domain, with the source and the filter being the pitch component and the envelope, respectively.

@@

Figure \ref{fig:deconvolution} shows an example of deconvolution of a CQT spectrogram into its envelope and pitch component. The CQT spectogram was computed from an audio signal created by concatenating 12 4-second notes of an acoustic bass playing from C1 (32.70 Hz) to B1 (61.74 Hz) in ascending pitch. The notes come from the NSynth dataset \cite{engel2017} and correspond to instrument id \texttt{bass\_acoustic\_000}, MIDI numbers \texttt{024} to \texttt{035}, and velocity number \texttt{075}. The CQT spectrogram was computed using librosa \cite{mcfee2015, schoerkhuber2010}, with a sampling rate of 16 kHz, a hop length of 512 samples, a minimum frequency of 32.70 Hz (corresponding to C1), 95 frequency bins, and 12 bins per octave. As we can see, the envelope looks as if the CQT spectrogram has been normalized in pitch, with all the notes being brought down to the lowest frequency (corresponding to C1); while the pitch component looks as if the CQT spectrogram has been stripped down from its energy, leaving mostly the fundamental frequencies of the notes. Note that in practice, we use a power CQT spectrogram (i.e., magnitude to the power of 2) and take the real part for the envelope and pitch component to ensure real values. We can also potentially post-process this deconvolution, for example, by zeroing the few negative values in the pitch component and using it to derive a refined  envelope from the log spectrum.

@@

Figure \ref{fig:extraction} shows an example of CQT-SECs (on the left plot). 20 coefficients were extracted from the envelope resulting from the deconvolution of the CQT spectrogram of the musical signal shown in Figure \ref{fig:deconvolution}. These coefficients essentially correspond to the harmonics of the musical instrument which contain most of its spectral energy and can therefore be a reasonable representation of its timbre. For comparison, we also show the MFCCs computed from the same musical signal (on the right plot). 20 coefficients were computed using librosa \cite{mcfee2015}, with a sampling rate of 16 kHz, a window length of 1024 samples, and hop length of 512 samples, matching the time resolution of the CQT-SECs so that both features have the same size in time and frequency.

\begin{figure*}[htp]
    \centering
    \includegraphics[width=\textwidth]{extraction.png}
    \caption{CQT-SECs extracted from the envelope obtained following the deconvolution of the CQT spectrogram shown in Figure \ref{fig:deconvolution} (left plot, shown in dB) and MFCCs computed from the same musical signal (right plot).}
    \label{fig:extraction}
\end{figure*}

@@

We derived the CQT-SECs for all the notes in this subset, by first computing the CQT spectrogram using librosa, with a sampling rate of 16 kHz, a hop length of 512 samples, a minimum frequency of 32.70 Hz (corresponding to C1), 95 frequency bins, and 12 bins per octave, and then extracting 20 coefficients from the envelope resulting from the deconvolution of the CQT spectrogram, leading to CQT-SECs of size 20 (coefficients) by 126 (time frames). We used a power CQT spectrogram (i.e., magnitude to the power of 2) and took the real part for the envelope to ensure real values. For comparison, we also computed the MFCCs, using librosa, with a sampling rate of 16 kHz, a window length of 1024 samples, a hop length of 512 samples, and 20 coefficients, leading to MFCCs of same size as the CQT-SECs, i.e., 20 by 126.

@@

We also averaged the note similarities over every one of the 11 instrument families, leading to similarity matrices of size 11 by 11, for both the CQT-SECs and the MFCCs. Figure \ref{fig:family_similarities} shows these instrument family similarity matrices and figure \ref{fig:family_similarities2} additionally shows the self-similarities (in green) and the cross-similarities (means and standard deviations in red) for every instrument family, for the CQT-SECs and the MFCCs. As we can again see, the similarities for the CQT-SECs have more variance compared with the MFCCs, and their self-similarities are noticeably and mostly higher than the means plus standard deviations of their cross-similarities while the self-similarities and cross-similarities for the MFCCs are all very high and harder to tell apart. We can clearly see with the CQT-SECs, for example, that the organ family has high self-similarity, showing that most of the instruments and notes in that family are quite similar to each other in terms of timbre. We can also see that the organ family is somewhat similar to the brass, flute, and vocal families but quite different from the other families. On the other hand, the string family has fairly low self-similarity, showing that either the instruments or notes in that family differ substantially from each other, or the CQT-SECs were not able to properly capture the overall timbre for this family.

@@

We have shown that we can design a simple but functional pitch-independent timbre feature which is well-adapted to musical data by decomposing the CQT spectrum into a pitch-independent spectral envelope and an energy-normalized pitch component, and extracting a number of coefficients from the spectral envelope. These CQT-SECs provide a compact and interpretable feature for characterizing the timbre of a musical instrument, showing higher variance in terms of similarities on a simple instrument similarity task, compared with the MFCCs which are commonly used by MIR practitioners to characterize timbre in music. For the interested reader, we provide a Python implementation of the CQT-SEC online with some examples\footnote{https://github.com/zafarrafii/CQT-SEC-Python}.
